---
title: "Lab 1"
author: "Hanna Fei"
format: html
editor: visual
embed-resources: true
fig-width: 8
fig-height: 8
fig-align: center
theme: sketchy
---

## Packages

```{r, message = FALSE}
library(datasauRus)
library(tidyverse)
```

### Question 1

```{r}
?datasaurus_dozen
table(datasaurus_dozen$dataset)
```

The datasaurus_dozen file has 1846 rows and 3 columns. It includes variables dataset, x, and y.

### Question 2

Plotting dino

```{r}
dino_data <- datasaurus_dozen[datasaurus_dozen$dataset == 'dino', ]
ggplot(data = dino_data, mapping = aes(x = x, y = y)) +
  geom_point()
```

Calculate correlation coefficient

```{r}
cor(dino_data$x, dino_data$y)
```

### Question 3

```{r}
star_data <- datasaurus_dozen[datasaurus_dozen$dataset == 'star', ]
ggplot(data = star_data, mapping = aes(x = x, y = y)) +
  geom_point()

cor(star_data$x, star_data$y)
```

The correlation coefficient for the star dataset is -0.062, very similar to the dino dataset, though slightly less extreme.

### Question 4

```{r}
circle_data <- datasaurus_dozen[datasaurus_dozen$dataset == 'circle', ]
ggplot(data = circle_data, mapping = aes(x = x, y = y)) +
  geom_point()

cor(circle_data$x, circle_data$y)
```

The r for circle is -0.068, slightly more extreme than the r for dino.

### Question 5

```{r}
#| fig-height: 10
#| fig-width: 10
par(mar = c(3,3,3,3))
layout(matrix(1:16, nrow=4, ncol=4))
for(name in unique(datasaurus_dozen$dataset)){
  subset <- datasaurus_dozen[datasaurus_dozen$dataset == name, ]
  plot(subset$x, subset$y, main = name)
}
layout(1)
```

### Question 6

```{r}
sapply(unique(datasaurus_dozen$dataset), function(name){
    subset <- datasaurus_dozen[datasaurus_dozen$dataset == name, ]
    return(cor(subset$x, subset$y))
})
```
